Diffantom: a software layer to build digital phantoms of whole-brain diffusion MRI

### 2015-07-14 paper/abbreviations.tex
finishing early version of paper
add details missing in paper
add new intro
towards new version of paper
A major update to the paper with new fig01 and coplete fig02

% -*- root: main.tex -*-
% @Author: Oscar Esteban
% @Date:   2015-06-16 15:45:32
% @Last Modified by:   Oscar Esteban
% @Last Modified time: 2015-06-17 16:28:24

\newacronym{dmri}{dMRI}{diffusion MRI}
\newacronym{hcp}{HCP}{Human Connectome Project}

>>>>

% -*- root: main.tex -*-
% @Author: Oscar Esteban
% @Date:   2015-06-16 15:45:32
% @Last Modified by:   Oscar Esteban
% @Last Modified time: 2015-07-14 12:49:57

\newacronym{dmri}{dMRI}{diffusion MRI}
\newacronym{fa}{FA}{fractional anisotropy}
\newacronym{adc}{ADC}{anisotropic diffusion coefficient}
% \newacronym{hcp}{HCP}{Human Connectome Project}
\newacronym{fod}{FOD}{fiber orientation distribution}
\newacronym{wm}{WM}{white matter}
\newacronym{gm}{GM}{gray matter}
\newacronym{cgm}{cGM}{cortical gray matter}
\newacronym{dgm}{dGM}{deep gray matter}
\newacronym{csf}{CSF}{cerebrospinal fluid}
\newacronym{snr}{SNR}{signal-to-noise ratio}
\newacronym{csd}{CSD}{constrained spherical deconvolution}
\newacronym{cc0}{CC0}{Creative Commons Zero licence}
\newacronym{roc}{ROC}{receiver operating characteristic}
\newacronym{pve}{PVE}{partial volume effect}
\newacronym{cst}{CST}{corticospinal tract}

\newglossaryentry{hcp}
{
	name={Human Connectome Project},
	text={HCP},
	first={Human Connectome Project (HCP, \cite{essen_human_2012})},
	long={Human Connectome Project (HCP)},
	description={Human Connectome Project}
}

### 2015-07-14 paper/main.tex
A major update to the paper with new fig01 and coplete fig02

\input{abbreviations.tex}
\begin{document}

@@

\noindent\textbf{\textit{Data Generation\textcolon}}
The methods to generate the data in this paper are inspired by the work of \cite{wilkins_fiber_2014} and
  described in \autoref{fig:figure1}.
\emph{Diffantom} integrates two major improvements with respect to \citep{wilkins_fiber_2014}.
First, since we use a dataset from the \gls*{hcp} as input, data are corrected for the most relevant distortions.
\cite{wilkins_fiber_2014} explicitly state that their original data were not subjected to certain distortion
  corrections, and thus, generated data are affected correspondingly.
Second improvement is a more advanced signal model to generate the phantom using
  \emph{phantomas} \citep{caruyer_phantomas_2014}.
By default, diffusion data are generated using a scheme of 100 directions distributed in two shells with uniform
  coverage \citep{caruyer_design_2013}.
Custom schemes can be generated supplying the tables of corresponding $b$-vectors and $b$-values.
Rician noise is also included in \emph{phantomas}, and can be set by the user.
The default value for \gls*{snr} is pre-set to 90.0.
\Gls*{csf} was simulated with a unique compartment with isotropic diffusivity $D_{CSF}$ of 3.0$\times$10$^{-6}$ [mm s$^{-1}$].
Conversely, \Gls*{wm} and \gls*{gm} included one isotropic compartment with $D_{WM} =$ and $D_{GM} =$, respectively.
Single fiber contributions had $D_{\parallel}=$ and $D_{\bot}=$.
All these values for diffusivity can be modified by the user for custom settings.

@@

\noindent\textbf{\textit{Conclusion\textcolon}}\label{sec:conclusion} %
\emph{Diffantom} is a software layer to be applied on \acrfull*{hcp} datasets, that is openly and freely distributed
  along with an example dataset.

\section*{Data Sharing}
The example dataset is available under the \gls*{cc0} in {\color{red} (dryad?)}.
Source code of \emph{diffantom} is available at \url{https://github.com/oesteban/phantomas} under an MIT licence.
\emph{Phantomas} is available in \url{https://github.com/ecaruyer/phantomas/} under the revised-BSD licence.

@@

\section*{Figures}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\linewidth]{figures/figure01}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure1} Figure \arabic{figure}. }%
{\emph{Diffantom} is a software layer wrapping the \acrfull{hcp} datasets to generate whole-brain diffusion \gls*{dmri} data.
First, the tissue fraction maps $\{T_1, T_2, T_3\}$ are computed using FSL FAST \citep{zhang_segmentation_2001}.
Second, the fiber orientation maps $\{V_1, V_2, V_3\}$ and corresponding volume fraction maps $\{F_1, F_2, F_3\}$ are
  obtained using the ball-and-stick model for multi-shell data of BEDPOSTX \citep{jbabdi_modelbased_2012}.
The third step takes the fiber orientations $\{V_1, V_2, V_3\}$ and their weights $\{F_1, F_2, F_3\}$ to compute the
  corresponding \glspl{fod} using spherical convolution of the single fiber response with each fiber orientation at
  each voxel.
The resulting \glspl{fod} map is then combined with three isotropic compartments corresponding to $\{T_1, T_2, T'_3\}$
  before generating the signal using \emph{phantomas} \citep{caruyer_phantomas_2014}.
The fraction map $T'_3 = T_3 - \sum F_i$ is the remaining isotropic fraction across the \gls*{wm} tissue map.
Finally, data can be used to compute the whole-brain tractography, for instance.}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\linewidth]{figures/figure02}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure2} Figure \arabic{figure}. }%
{\textbf{A. Recommended use of \emph{diffantom}}.
The dataset can be used to test algorithms and pipelines against a perturbation that is
  introduced in \emph{diffantom}.
Perturbations can model typical artifacts found in \gls*{dmri} datasets or pathological conditions
  such as a tumor.
\textbf{B1, B2, B3. Tractograms of fundamental fiber bundles}.
}
\end{figure}

>>>>

\input{abbreviations.tex}

\newcommand{\e}[1]{\ensuremath{\;\cdot\,}10\ensuremath{^{#1}}}
\begin{document}

@@

\noindent\textbf{\textit{Data Generation\textcolon}\label{sec:data_generation}}
The methods to generate the data in this paper are inspired by the work of \cite{wilkins_fiber_2014} and
  described in \autoref{fig:figure1}.
\emph{Diffantom} integrates two major improvements with respect to \citep{wilkins_fiber_2014}.
First, since we use a dataset from the \gls*{hcp} as input, data are corrected for the most relevant distortions.
\cite{wilkins_fiber_2014} explicitly state that their original data were not subjected to certain distortion
  corrections, and thus, generated data are affected correspondingly.
Second improvement is a more advanced signal model to generate the phantom using
  \emph{phantomas} \citep{caruyer_phantomas_2014}.

In order to generate \emph{a diffantom}, the simulation inputs (see \autoref{fig:figure1}, A) are computed.
On one hand, the software requires up to five volume fractions $\{T_{1\,\cdots\,5}\}$ corresponding to the isotropic
  compartments that will be simulated.
The isotropic fraction maps use the \emph{5TT format} established with \emph{MRTrix3} \citep{tournier_mrtrix_2012}
  that describes the structure of the brain in the following order: \gls*{cgm}, \gls*{dgm}, \gls*{wm}, \gls*{csf},
  and abnormal tissue.
Since we simulated a healthy subject, the last fraction map is omitted.
On the other hand, the anisotropic compartments are specified by up to three volume fractions $\{F_{1\,\cdots\,3}\}$
  of three single fiber populations per voxel along with their corresponding direction maps $\{V_{1\,\cdots\,3}\}$.
The process to obtain these inputs is described as follows:
1) The fiber orientation maps $\{V_{1\,\cdots\,3}\}$ and their corresponding single-fiber fraction maps are
  obtained using the ball-and-stick model for multi-shell data of BEDPOSTX \citep{jbabdi_modelbased_2012}
  on the \gls*{dmri} data of a certain \gls*{hcp} subject.
2) We obtain a \gls*{fa} map after fitting a tensor model with \emph{MRTrix3}.
3) The original single-fiber fractions and the \gls*{fa} map are smoothed with a non-local means filter included
  in \emph{dipy} \citep{garyfallis_dipy_2011}.
4) By combining with \emph{MRTrix} the volume fraction maps computed with FSL FAST \citep{zhang_segmentation_2001} and
  FSL FIRST \citep{patenaude_bayesian_2011}, we obtain the macro-structural 5TT map of the subject.
5) The objects obtained in the previous steps are combined as described in the \nameref{sec:appendix} to generate the
  outcome presented in \autoref{fig:figure1}, box A.

Once a micro-structural model of the volume has been synthesized, a map of \glspl{fod} is computed using spherical
  convolution of the single fiber response and the fiber orientation maps $\{V_{1\,\cdots\,3}\}$, weighted by the
  corresponding single-fiber fractions $\{F_{1\,\cdots\,3}\}$.
The single fiber response is {\color{red} (Manu, please help me on define this)}, with $D_{\parallel}=$ 2.2\e{-3}
  and $D_{\bot}=$ 0.2\e{-3} [mm s$^{-1}$].
The resulting \glspl{fod} map is then combined with the isotropic compartments corresponding to $\{T_{1\,\cdots\,5}\}$
  before generating the signal using \emph{phantomas} \citep{caruyer_phantomas_2014}.
By default, diffusion data are generated using a scheme of 100 directions distributed in one shell with uniform
  coverage \citep{caruyer_design_2013}.
Custom one- or multi-shell schemes can be generated supplying the tables of corresponding $b$-vectors and $b$-values.
Rician noise is also included in \emph{phantomas}, and can be set by the user.
The default value for \gls*{snr} is pre-set to 90.0.
\Gls*{csf} was simulated with a unique compartment with isotropic diffusivity $D_{CSF}$ of 3.0\e{-3} [mm s$^{-1}$].
Conversely, \Gls*{wm} and \gls*{gm} included one isotropic compartment with $D_{WM} =$ 2.0\e{-4}, $D_{cGM} =$ 7.0\e{-4}
  and $D_{dGM} =$ 9.0\e{-4}, respectively [mm s$^{-1}$].
All these values for diffusivity (and the corresponding to the single-fiber response) can be modified by the user with
  custom settings.

@@

\noindent\textbf{\textit{Conclusion\textcolon}}\label{sec:conclusion} %
\emph{Diffantom} is a digital phantom generated from a dataset belonging to the \acrlong*{hcp}, and the software layer to
  simulate it.
A first \emph{diffantom} is presented here to be openly and freely distributed along with the necessary software to reproduce
  this dataset and generate new \emph{diffantoms}.

\section*{Data Sharing}
The first \emph{diffantom} is available under the \gls*{cc0} in {\color{red} (dryad?)}.
The associated software to ``\emph{diffantomize}'' real \gls*{dmri} data is available at \url{https://github.com/oesteban/diffantom}
  under an MIT licence.
\emph{Phantomas} is available in \url{https://github.com/ecaruyer/phantomas} under the revised-BSD licence.

@@

\section*{Appendix}\label{sec:appendix}
Let $\{T'_{1\,\cdots\,5}\}$ be the set of original fractions maps obtained with the combined use of FAST and FIRST, FA the
  \gls*{fa} map obtained from the original \gls*{dmri} data, and $\{V_{1\,\cdots\,3}, F'_{1\,\cdots\,3}\}$ the direction
  maps $V_i$ of single-fibers with their estimated volume fractions $F'_i$ calculated with BEDPOSTX.
The final $\{T_{1\,\cdots\,5}\}$ maps of isotropic fractions are computed as follows:

  \begin{align*}
  T_1 &= (1.0-w_{cgm}) \cdot T'_1 \\
  T_2 &= (1.0-w_{dgm}) \cdot T'_2 \\
  T_3 &= (1.0-w_{wm}) \cdot T'_3 \\
  T_4 &= T'_4 \\
  T_5 &= 0.0
  \end{align*}
where $w_{\{cgm, dgm, wm\}}$ are weighing factors of anisotropic diffusion for each tissue
  (defaults are $w_{cgm} =$ 25\%, $w_{dgm} =$ 50\%, $w_{cgm} =$ 95\%).
The final $\{F_{1\,\cdots\,3}\}$ maps are computed as follows:

\begin{align*}
F_1 &= w_{wm} \cdot T_2 \cdot \text{FA} + w_{f1} (w_{cgm} \cdot T_1 + w_{dgm} \cdot T_2) \\
F_2 &= w_{wm} \cdot T_2 - (F_1 + F_3) + w_{f2} (w_{cgm} \cdot T_1 + w_{dgm} \cdot T_2) \\
F_3 &= w_{wm} \cdot F'_3 + w_{f3} (w_{cgm} \cdot T_1 + w_{dgm} \cdot T_2)
\end{align*}
where $w_{\{f1, f2, f3\}}$ are the contributions of the \gls*{gm} compartments to each single-fiber population.
By default: $w_{f1} = $ 48\%, $w_{f2} = $ 37\%, $w_{f3} = $ 15\%.
Finally, the resulting maps are normalized to fulfill $\sum_i T_i + \sum_i F_i = 1.0$.

\newpage
\section*{Figures}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\linewidth]{figures/figure01A}
\includegraphics[width=\linewidth]{figures/figure01B}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure1} Figure \arabic{figure}. }%
{\textbf{Generation of \emph{diffantoms}:} First, a micro-structural model is generated as described
  in the \nameref{sec:data_generation} subsection.
This micro-structural model is feeded into a simulation block that generates the final \gls*{dmri} dataset
  as described in the second block.
}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\linewidth]{figures/figure02}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure2} Figure \arabic{figure}. }%
{\textbf{A. Recommended use of \emph{diffantom}}.
The dataset can be used to test algorithms and pipelines against a perturbation that is
  introduced in \emph{diffantom}.
Perturbations can model typical artifacts found in \gls*{dmri} datasets or pathological conditions
  such as a tumor.
\textbf{B1. Tractogram of fibers crossing slice 56 of the \emph{diffantom}} represented over that corresponding
  slice of the \emph{b0} volume. \textbf{B2. Segmentation of some fiber bundles}: in blue color is represented
  the right \gls*{cst}, in red the left \gls*{cst}, and in green the forceps minor.
  The box includes the slice 56 of the \emph{b0} and the light gray shadow is the pial surface.
}
\end{figure}
